Wait-free Queues with Polygarithmic Step Complexity

### 2022-09-29 backup/jul 14.tex
fix citations, review index deq

\usepackage[b4paper,left=0.8in,right=0.8in,top=0.8in,bottom=0.8in]{geometry}
\usepackage{tikz-qtree}
\usepackage{algorithm}
\usepackage{algpseudocode}
\makeatletter
\renewcommand{\ALG@beginalgorithmic}{\footnotesize}
\makeatother
\usepackage{graphicx}
\usepackage{subcaption}
\usepackage{showkeys}
\usepackage{amsmath}

@@

\renewcommand\thealgorithm{}
\newcommand{\setalglineno}[1]{%
  \setcounter{ALC@line}{\numexpr#1-1}}

@@

\usepackage{amsmath,amssymb,amsthm}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}[theorem]{Lemma}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}

@@

\paragraph{}
We choose to linearize operations in a block from the left child before those from the right child as a convention. Operations within a block of the root can be ordered in any way that is convenient. In effect, this means that if there are concurrent new blocks in a \textsc{Refresh}() step from several processes we linearize them in the order of their process ids. So for example  operations aggregated in block (10,2) are in the order (2,3),(4,1),(0,2). All blocks from the left child with come before the right child and the order of blocks of each child is preserved among themselves.
%\begin{lemma}
%\label{lem:block_size}
%There cannot be more than one operation from each process in a \textsc{Refresh}() invocation.
%\end{lemma}
%\begin{proof}
%  If so, then that parent process has invoked two simultaneous operations. $ \blacksquare $
%\end{proof}

@@

\subsection{Block Tree Algorithm}
\paragraph{}
Our Block Tree is a linearizable implementation of a data structure that stores a sequence of elements. It has two methods (see Algorithm~\ref{alg2}), \textsc{Append}($e$) which appends element $e$ to the sequence, and \textsc{Get}($i$) which returns the $i$th element in the sequence.

@@

Each process is assigned to a leaf in a shared tournament tree. Thus, for example, the leaf node for process $p_i$ contains an array of elements by $p_i$ in the order they were invoked.
Each internal node of the tree contains an array of blocks of elements.
Block $b$ in node $n$ is created in a \textsc{Propagate}() step and is merged block of new blocks at the time of \textsc{Propagate}() reading $n$'s children blocks. Each block consists of pointers left and right, to the last block merged into itself from left and right child in that order. Moreover, two numbers, left and right, indicate the count of elements in the blocks from the left and right child consecutively. Furthermore, prefix left, and right can be computed from the prefix sum of left and right values.
Elements of block $b$ can be determined recursively (\textsc{GetElements($b$)}).
The $i$th element in the sequence can be determined in \textsc{O}($\log^2 p$) steps by recursively finding $i$th element in block $b$ (\textsc{GetElement($i$)})
After element $e$ is propagated (appended to a block int the root), its index can be computed with \textsc{GetIndex}($op$).

@@

\paragraph{Specification}
A block tree is a shared data structure that stores a sequence of elements.  It has two methods \texttt{Append(e)} and \texttt{Get(i)}. \texttt{Append(e)} adds \texttt{e} to the end of the sequence and returns the index of \texttt{e} in the sequence. \texttt{Get(i)} returns \texttt{i}th element stored in the sequence.

\paragraph{SubBlock} Block \texttt{s} is a subblock of \texttt{b} if \texttt{s} is between blocks \texttt{start..end} in \texttt{n} from Lines 41,42 of \texttt{CreateBlock()}.

\paragraph{Membership}
Element \texttt{e} is a member of block \texttt{b} in:
\begin{itemize}
  \item internal node \texttt{n}, if \texttt{e} is a memeber of \texttt{s} that \texttt{s} is a subblock of \texttt{b}.
  \item leaf node \texttt{n}, if \texttt{e} belongs to \texttt{n.dir.blocks[b\textsuperscript{$\prime$}.end\textsubscript{dir}+1..b.end\textsubscript{dir}]} for \texttt{dir} $\in$ \texttt{\{left, right\}} which b\textsuperscript{$\prime$} is the previous block of \texttt{b} in \texttt{n}.
\end{itemize}

\paragraph{Order of elements inside node}
Element \texttt{d} is before element \texttt{e} in node \texttt{n}, if:
\begin{itemize}
  \item The block containing d is before the block containing e.
  \item e and d are in the same block and d is in the left child and e is in the right child.
  \item d is before e in the same child's order.
\end{itemize}

\begin{figure}[hbt]
  \center\includegraphics[width=5.5in]{pics/tree}
  \caption{Order of elements in b: elements in leaves are ordered with numerical order in the drawing.}
\end{figure}

\paragraph{\texttt{CreateBlock()}} \texttt{CreateBlock(n)} returns a block containing new operations of \texttt{n}'s children. \texttt{b\textsuperscript{$\prime$}.end\textsubscript{left}} stores the index of the rightmost subblock of left child of \texttt{b}'s previous block. Other attributes are assigned values followed by definition.
\begin{figure}[hbt]
  \center\includegraphics[width=5.5in]{pics/createblock}
  \caption{\label{fig::createBlock}Snapshot of a CreateBlock()}
\end{figure}

%\textsf{Is it necessary to say \texttt{CreateBlock()} is doing what it's supposed to do by definition? Or \texttt{Append()} for example?}

\paragraph{Double Refresh}
Elements in \texttt{n}'s children's blocks before line \texttt{13} are guaranteed to be in \texttt{n}'s blocks after Line~\texttt{15}.
\begin{proof}
\texttt{CreateBlock()} reads blocks in the children that does not exist in the parent and aggregates them into one block. If a \texttt{Refresh} procedure returns true it means it has appended the block created by \texttt{CreateBlock()} into the parent node's sequence. So suppose two \texttt{Refresh}es fail. Since the first \texttt{Refresh} was not successful, it means another CAS operation by a \texttt{Refresh}, concurrent to the first\texttt{Refresh}, was successful before the second \texttt{Refresh}. So it means the second failed \texttt{Refresh} is concurrent with a successful \texttt{Refresh} that assuredly has read block before the mentioned line \texttt{13}. After all it means if any of the \texttt{Refresh} attempts were successful the claim is true, and also if both fail the mentioned claim still holds.
\end{proof}
\begin{figure}[hbt]
  \center\includegraphics[width=4in]{pics/doublerefresh.png}
  \caption{The second failed Refresh is assuredly concurrent to a Successful \texttt{Refresh()} with \texttt{CreateBlock} line after first failed \texttt{Refresh}'s \texttt{CreateBlock()}.}
\end{figure}

\paragraph{Disjunction} Blocks in node \texttt{n}'s contain disjoint sets of elements.
\begin{proof}
Without loss of generality, assume blocks \texttt{b1, b2} contain common element \texttt{e} from the left child, and \texttt{b2} is after \texttt{b1} in \texttt{n}'s sequence of blocks. So block start of b2's \texttt{CreateBlock()} is after block end of \texttt{b1}'s end. Since \texttt{b2}'s start is the end of the block before itself, it cannot be before \texttt{b1}'s end.
\end{proof}

\paragraph{Total Order}
Sequence represented by the Block Tree is the sequence of the blocks stored in the root.

\paragraph{Linearization Points}
\texttt{Get(i)} is linearized when it terminates. \texttt{Append(e)} is linearized right after when a block containing \texttt{e} is appended to the root, if there are multiple elements appended together, they are linearized by the defined order in the root.

\paragraph{Subblocks Upperbound} Block b has at most $p$ subblocks.
\begin{proof}
If there are more than $p$ subblocks, then there is more than one block from process pl. \texttt{Append(e)} finishes after propagating and appending \texttt{e} to the root(line 9). So these blocks cannot be appended to root already, so pl has invoked two concurrent \texttt{Append()}s(line 1) without terminating the first one.
\end{proof}

\paragraph{Computing \texttt{Get(n, b, i)}}
To find the \texttt{i}th element in block \texttt{b} of node \texttt{n}, we search among subblocks of \texttt{b} that is bounded by $p$. Subblocks of a block are within the start and end block of the \texttt{CreateBlock()} procedure of it.

\paragraph{How \texttt{Refresh(n)} works.}
\begin{enumerate}
  \item Read n's counter and head
  \item Create block b
  \item CAS b into n
  \item If previous succeed:
  \begin{enumerate}
  \item Update sup of b's ending subblocks
  \item Increment children's counters
  \end{enumerate}
\end{enumerate}

%\paragraph{Time-Space Relation} There are at most $p^2$ blocks with same \texttt{time} attribute.
%\begin{proof}
%If a process dies at line 28, the children counter won't increase, and each process aggregates most block, and since there are at most p concurrent processes, then there will be at most $p^2$ blocks with same \texttt{time} attribute.
%\end{proof}
%
%\paragraph{Time-Propagate() Relation} In a \texttt{CreateBlock()} procedure, the blocks that are seen have at most 2 different \texttt{time} attribute values.
%
%\paragraph{Time-Super Relation} sup[i+1]-sup[i] is $\leq p$
%
%\paragraph{\texttt{b}'s Superblock position} SuperBlock of \texttt{b} is within range $\pm p$ of the \texttt{\texttt{sup[b.time]}}.
%\begin{proof}
%
%\end{proof}
\begin{figure}[hbt]
  \center\includegraphics[width=6in]{pics/super}
  \caption{\texttt{Sup} and \texttt{timer} in a node, numbers on blocks are their \texttt{time} values.}
\end{figure}

\paragraph{Computing superblock}
\begin{enumerate}
  \item Value read for \texttt{super[b.time]} in line 71 is not null.\begin{proof}
    \texttt{Index()} is invoked after finishing \texttt{Propagate()} in line 10. For each value \texttt{c\textsubscript{dir}} read in lines 23, \texttt{super} is set before incrementing in lines 26,27.
  \end{proof} 
  \item \texttt{super[]} preserves order from child to parent; if in a child block \texttt{b} is before \texttt{c} then \texttt{b.time} $\leq$ \texttt{c.time} and \texttt{super[b.time]} $\leq$ \texttt{super[c.time]}\begin{proof}
   Follows from the order of lines 37, 26, 27.
 \end{proof}
%  \item In a propagate step at most 2 different time values are read \\ If there are more than 2 numbers then the smallest number should have been propagated far before.
%  \item There are at most $p^2$ blocks with same time value in a node. \\ At most p processes could die before line 27 and each contains at most p elements.
  \item \texttt{super[i+1]-super[i]}$\leq p$
\begin{proof}
 In a Refresh with successful CAS in line 24, \texttt{super} and \texttt{counter} are set for each child in lines 26,27. Assume the current value of the counter in node \texttt{n} is \texttt{i+1} and still \texttt{super[i+1]} is not set. If an instance of successful \texttt{Refresh(n)} finishes \texttt{super[i+1]} is set a new value and a block is added after \texttt{n.parent[sup[i]]}. There could be at most $p$ successful unfinished concurrent instances of \texttt{Refresh()} that have not reached line 27. So the distance between \texttt{super[i+1]} and \texttt{super[i]} is less than $p$.
 \end{proof}
  \item Superblock of \texttt{b} is within range $\pm 2p$ of the \texttt{super[b.time]}.
\begin{proof}
\texttt{super[i]} is the index of the superblock of a block containing block b. It is trivial to see that \texttt{n.super} and \texttt{n.b.counter} are increasing.  \texttt{super(b)} is the real superblock of b. \texttt{super(t]} is the index of the superblock of the last block with time \texttt{t}. If \texttt{b.time} is \texttt{t} we have:
$$super[t]-p\leq super[t-1]\leq super(t-1] \leq super(b) \leq super(t+1)\leq super(t+1]\leq super[t]+p$$
  \end{proof} 
\end{enumerate}

%The block tree contains a tournament tree in which leaves are assigned to processes, and each process appends elements to its leaf and tries to propagate it up to the root. The final sequence is stored at the root. \texttt{Append} after appending to leaf calls \texttt{Propagate} which iterates over the path to root. In each \texttt{Propagate(n)} invocation on node n it tries ti \texttt{Refresh} the sequence stored at node n with the new elements in n's children. After adding the element to the root \texttt{Index(op)} finds the order of the element in the root.

%\textsf{I finally thought maybe a sequence of elements is better for telling specification.}

@@

\Statex $\blacktriangleright$ \texttt{\textsl{element} e}
\begin{itemize}
%\item \texttt{\textsl{string} name}
\item \texttt{\textsl{int} pid}
\item \texttt{\textsl{int} loc\textsf{: location in \texttt{l\textsubscript{pid}}.ops}}
\end{itemize}

\Statex $\blacktriangleright$ \texttt{\textsl{node} n}
\begin{itemize}
\item \texttt{\textsl{*node} left, right, parent}
\item \texttt{\textsl{block[]} blocks}\textsf{: blocks stored in \texttt{n}}
\item \texttt{\textsl{int} head= 1}\textsf{: index of first empty cell of \texttt{blocks}}
\item \texttt{\textsl{int} counter= 0}\textsf{}
\item \texttt{\textsl{int[]} super}\textsf{: \texttt{super[i]} stores the index of a superblock in parent that contains some block of this node whose \texttt{time} is \texttt{i}}
\end{itemize}

\Statex $\blacktriangleright$ \texttt{\textsl{leaf node} l\textsubscript{i} extends \textsl{node}}
\begin{itemize}
\item \texttt{\textsl{operation[]} ops}\textsf{: elements that are invoked to \texttt{Append()} to the blok tree by process i}
\end{itemize}

\Statex $\blacktriangleright$ \texttt{\textsl{block} b}
\begin{itemize}
  \item \texttt{\textsl{int} num\textsubscript{left}, sum\textsubscript{left}}

  \textsf{~~\#operations from the left subblocks of b, prefix sum of num\textsubscript{left}}

  \item \texttt{\textsl{int} num\textsubscript{right}, sum\textsubscript{right}}

  \textsf{~\#operations from the right subblocks of b, prefix sum of num\textsubscript{right}}

  \item \texttt{\textsl{int} sum}

  \textsf{\# operation in b}

  \item \texttt{\textsl{int} end\textsubscript{left}, end\textsubscript{right}}

  \textsf{~~index of \texttt{b}'s last subblock}
  \item \texttt{\textsl{int} time}

\end{itemize}

%\Statex \textbf{Examples}
%\Statex \hspace{0pt} \texttt{ l\textsubscript{i}[j]}\textsf{: jth operation of ith process}
%\Statex \hspace{0pt} \texttt{ n.left}\textsf{: left child of node n}
%\Statex \hspace{0pt} \texttt{ n.blocks[n.head]}\textsf{: leftmost empty block of node n}
%\Statex \hspace{0pt} \texttt{ n.blocks[i].num\textsubscript{left}}\textsf{: \#left operations of ith block of n}
%\Statex \hspace{4pt} \texttt{b.end\textsubscript{right}}\textsf{: index of ending right subblock of b}

@@

\Function{int}{Append}{\textsl{operation} op, \textsl{int} pid} 
\State \texttt{op.loc= l\textsubscript{pid}.head}
\State \texttt{block b= \Call{new}{\textsl{block}}}
\State \texttt{b.time= op.loc}
\State \texttt{b.sum= 1}
\State \texttt{l\textsubscript{pid}.ops[op.loc]= op}
\State \texttt{l\textsubscript{pid}.blocks[op.loc]= b}
\State \texttt{op.head+= 1}
\State \Call{Propagate}{l\textsubscript{pid}.parent}
\State \Return \Call{Index}{l\textsubscript{pid}, op.loc, b} 
\EndFunction{Append}

@@

\Function{void}{Propagate}{\textsl{node} n}
\If{\texttt{not} \Call{Refresh}{n}}
\State \Call{Refresh}{n}
\EndIf
\If{\texttt{n {\keywordfont is} root}} \Return
\EndIf
\State \Call{Propagate}{n.parent}
\EndFunction{Propagate}

@@

\Function{boolean}{Refresh}{\textsl{node} n}
\State \texttt{i= n.head}
\State \texttt{c= n.counter}
\State \texttt{new, c\textsubscript{left}, c\textsubscript{right}= \Call{CreateBlock}{n, i+1, c}}
\If{\Call{CAS}{n.blocks[i+1], null, new}}
\ForEach{\texttt{dir} {\keywordfont{in}} \texttt{\{left, right\}}}
\State \texttt{\Call{CAS}{n.dir.super[c\textsubscript{dir}], null, i+1}}
\State \texttt{\Call{CAS}{n.dir.counter, c\textsubscript{dir}, c\textsubscript{dir}+1}}
\EndFor
\State \Return{\texttt{true}}
\Else 
\State \Return{ \texttt{false}}
\EndIf
\State \texttt{\Call{CAS}{n.head, i, i+1}}
\EndFunction{Index}

@@

\Function{block, int, int}{CreateBlock}{\textsl{node} n, \textsl{int} i, int c} 
\Statex\Comment{Creates a block to insert into \texttt{n.blocks[i]}}
\State \texttt{block b= \Call{new}{\textsl{block}}}
\State \texttt{b.time= c}
\ForEach{\texttt{dir} {\keywordfont{in}} \texttt{\{left, right\}}}
\State \texttt{j= n.dir.head}
\State \texttt{start= n.blocks[i-1].end\textsubscript{dir}}
\State \texttt{end= n.dir.blocks[j]}
\State \texttt{c\textsubscript{dir}= n.dir.blocks[j].time}
\State \texttt{b.end\textsubscript{dir}= j}
\State \texttt{b.num\textsubscript{dir}= end.sum - start.sum}
\State \texttt{b.sum\textsubscript{\#dir}= n.blocks[i-1].sum\textsubscript{\#dir} + b.num\textsubscript{dir}}
\EndFor
\State \texttt{b.sum= b.sum\textsubscript{left} + b.sum\textsubscript{right}}
\State \Return \texttt{b, c\textsubscript{left}, c\textsubscript{rightt}}
\EndFunction{CreateBlock}

@@

\begin{algorithm}
\caption{Block Tree Continued}
\begin{algorithmic}[1]
\setcounter{ALG@line}{46}

\Function{element}{Get}{\textsl{int} i} 
\State \texttt{res= \Call{BSearch}{root, sum, i, 0, root.head}}
\State \Return{\Call{Get}{root, res, i-root.blocks[res-1].sum}}
\EndFunction{Get}

@@

\Statex $\leadsto$ \textsf{Precondition: \texttt{n.blocks[start..end]} contains a block with field \texttt{f} $\geq$ \texttt{i}}
\Function{int}{BSearch}{\textsl{node} n, \textsl{field} f, \textsl{int} i, \textsl{int} start, \textsl{int} end}

@@

\Function{element}{Get}{\textsl{node} n, \textsl{int} b, \textsl{int} i} 
\Comment{\textmd{Returns the \texttt{i}th operation in \texttt{b}th block of node \texttt{n}}}
\If{\texttt{n {\keywordfont is} leaf}} \Return \texttt{n.ops[i]}
\Else
\If{\texttt{i $\leq$ n.blocks[b].num\textsubscript{left}}} \Comment{\texttt{i} exists in left child of \texttt{n}}
\State \texttt{subBlock= \Call{BSearch}{n.left.sum, i, n.blocks[b-1].end\textsubscript{left}+1, n.blocks[b].end\textsubscript{left}}}
\State \Return\Call{Get}{n.left, subBlock, i-n.left.blocks[subBlock-1].sum} 
\Else
\State \texttt{i= i-n.blocks[b].num\textsubscript{left}}
\State\texttt{subBlock=\Call{BSearch}{n.right.sum, i, n.blocks[b-1].end\textsubscript{right}, n.blocks[b].end\textsubscript{right}}}
\State \Return\Call{Get}{n.right, subBlock, i-n.right.blocks[subBlock-1].sum} 
\EndIf
\EndIf
\EndFunction{Get}

\Statex
\Statex $\leadsto$ \textsf{Precondition:~\texttt{i}th operation in node \texttt{n} is in block \texttt{b} of node \texttt{n}.}
\Function{index}{Index}{\textsl{node} n, \textsl{int} i, \textsl{int} b} \Comment{Returns rank in the  
root of the \texttt{i}th operation in node \texttt{n}.}
\If{\texttt{n {\keywordfont is} root}} \Return \texttt{i}
\Else
\State \texttt{dir= (n.parent.left==n)? left: right}
\State \texttt{superBlock= \Call{BSearch}{n.parent, n.sum\textsubscript{dir}, i, sup[n.blocks[b].time]-p, sup[n.blocks[b].time]+p}}
\If{\texttt{dir {\keywordfont is} left}}
\State \texttt{i+= n.parent.blocks[superBlock-1].sum\textsubscript{right}}
\Else \State \texttt{i+= n.parent.blocks[superBlock].sum\textsubscript{left} + n.parent.blocks[superBlock].sum\textsubscript{left} + n.blocks[nBlock-1].sum}
\EndIf
\State \Return\Call{Index}{n.parent, i, superBlock}
\EndIf
\EndFunction{Index}

@@

\section{Implementing Queue using Block Tree}
\subsection{Idea in a nutshell}
With the ideas introduced in block tree we are going to create a shared wait-free queue with \textsc{O}$(\log ^2p+\log n)$ steps. A queue stores a sequence of elements and supports two operations, enqueue and dequeue. \texttt{Enqueue(e)} appends element \texttt{e} to the sequence stored.\texttt{Dequeue()} removes and returns the first element among in the sequence. If the queue is empty it returns \texttt{null}. Knowing index $i$ is the tail of the queue, we can return the dequeue response using \texttt{Get(i)}.  So in the rest we modify block tree to compute \texttt{i} for each \texttt{Dequeue()} to achieve a FIFO queue.
%
%\subsection{What should a \texttt{Dequeue()} return, regarding history of operations?}
\paragraph{}
Next, we describe how to use block tree to implement queues. The block tree, maintains the history of all operations, not only the current state of the queue. Now consider the following history of operations. What should each \texttt{Dequeue()} return?

@@

\begin{definition}
A non-null dequeue is one that returns a non-null value.
\end{definition}

\paragraph{}
In the example above, \texttt{Dequeue()} operations return \texttt{5, 2, 3, null, 4} in order. Before \texttt{ENQ(4)} the queue gets empty so the last \texttt{DEQ()} returns null. If the queue is non-empty and $r$ \texttt{Dequeue()} operations have returned a non-null response, then $i$th \texttt{Dequeue()} returns the input of the $r+1$th \texttt{Enqueue()}. So, in order to answer a Dequeue, it's sufficent to know the size of the queue and the number of previous non-null dequeues.

%\texttt{DEQ[i] = (size>0) ? ENQ[r+1] : null;}

%\subsection{What should a \texttt{Dequeue()} return, regarding history of blocks of operations?}

>>>>

\usepackage[a4paper,left=0.8in,right=0.8in,top=1.1in,bottom=1.1in]{geometry}
\usepackage{tikz-qtree}
\usepackage{algorithm}
\usepackage{algpseudocode}
\makeatletter
\renewcommand{\ALG@beginalgorithmic}{\footnotesize}
\makeatother
\usepackage{graphicx}
\usepackage{subcaption}

@@

@@

\usepackage{amsmath,amssymb,amsthm}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}[theorem]{Lemma}
\theoremstyle{definition}
\newtheorem{definition}{Definition}

@@

\paragraph{}
We choose to linearize operations from the left child before those from the right child as a convention. In effect, this means that if there are concurrent operations in a \textsc{Refresh}() step from several processes we linearize them in the order of their process ids. So for example  operations aggregated in block (10,2) are in the order (2,3),(4,1),(0,2). All blocks from the left child with come before the right child and the order of blocks of each child is preserved among themselves.
%\begin{lemma}
%\label{lem:block_size}
%There cannot be more than one operation from each process in a \textsc{Refresh}() invocation.
%\end{lemma}
%\begin{proof}
%  If so, then that parent process has invoked two simultaneous operations. $ \blacksquare $
%\end{proof}

@@

\subsection{Block Tree Algorithm}
\paragraph{}
Our Block Tree is a linearizable implementation of a data structure that stores the elements appended by processes. It has two methods (see Algorithm~\ref{alg2}), \textsc{Append}($e$) which appends element $e$ to the sequence, and \textsc{GetIndex}($i$) which returns the $i$th element in the sequence.

@@

Each process is assigned with a leaf in a shared tournament tree. Thus, for example, the leaf node for process $p_i$ contains an array of operations by $p_i$ in the order they were invoked.
Each internal node of the tree contains an array of blocks of elements.
Block $b$ in node $n$ is created in a \textsc{Propagate}() step and is merged block of new blocks at the time of \textsc{Propagate}() reading $n$'s children blocks. Each block consists of pointers left and right, to the last block merged into itself from left and right child in that order. Moreover, two numbers, left, right indicating the count of elements in the blocks from left and right child consecutively. Furthermore, two amounts, prefix left and prefix right, can be computed from the prefix sum of left and right values.
Elements of block $b$ can be determined recursively (\textsc{GetElements($b$)}).
Element $i$th in the sequence can be determined in \textsc{O}($\log^2 p$) steps by recursively finding $i$th element in block $b$ (\textsc{GetElement($i$)})
After element $e$ is propagated (appended to a block int the root), its index can be computed with \textsc{GetIndex}($op$).

@@

@@

\Statex $\blacktriangleright$ \texttt{\textsl{operation} op}
\begin{itemize}
%\item \texttt{\textsl{string} name}
\item \texttt{\textsl{int} pid}
\item \texttt{\textsl{int} loc\textsf{: location in \texttt{l\textsubscript{pid}}'s array}}
\end{itemize}

\Statex $\blacktriangleright$ \texttt{\textsl{leaf} l\textsubscript{i}}
\begin{itemize}
\item \texttt{\textsl{*node} parent}\textsf{: pointer to parent in Tree}
\item \texttt{\textsl{operation[]} ops}\textsf{: ops invoked by process i}
\item \texttt{\textsl{int} last= 1}\textsf{: index of first empty cell of \texttt{ops}}
\end{itemize}

\Statex $\blacktriangleright$ \texttt{\textsl{node} n}
\begin{itemize}
\item \texttt{\textsl{*node} left, right, parent}
\item \texttt{\textsl{block[]} blocks}\textsf{: blocks propagated to \texttt{n}}
\item \texttt{\textsl{int} last= 1}\textsf{: index of first empty cell of \texttt{blocks}}
\item \texttt{\textsl{int} counter= 0}\textsf{: \#groups of blocks propagated from \texttt{n}}
\item \texttt{\textsl{int[]} super}\textsf{: approximate index of superblock of a block in the given group in \texttt{n}'s parent}
\end{itemize}

\Statex $\blacktriangleright$ \texttt{\textsl{block} b}
\begin{itemize}
  \item \texttt{\textsl{int} num\textsubscript{left}, num\textsubscript{right}, sum, sum\textsubscript{left}, sum\textsubscript{right}}\textsf{: \#operations from the left and the right child, sum of all operations, prefix sum of operations from the left and the right child}
  \item \texttt{\textsl{int} end\textsubscript{left}, end\textsubscript{right}}\textsf{: index of \texttt{b}'s last subblock}
  \item \texttt{\textsl{int} group}\textsf{: approximate group number}
\end{itemize}

\Statex \textbf{Examples}
\Statex \hspace{0pt} \texttt{ l\textsubscript{i}[j]}\textsf{: jth operation of ith process}
\Statex \hspace{0pt} \texttt{ n.left}\textsf{: left child of node n}
\Statex \hspace{0pt} \texttt{ n.blocks[n.last]}\textsf{: index of last block of node n}
\Statex \hspace{0pt} \texttt{ n.blocks[i].num\textsubscript{left}}\textsf{: \#left operations of ith block of n}
\Statex \hspace{4pt} \texttt{b.end\textsubscript{right}}\textsf{: index of ending right subblock of b}

@@

\Function{int}{Append}{\textsl{operation} op, \textsl{int} pid} 
\State \texttt{op.loc= l\textsubscript{pid}.last + 1}
\State \texttt{l\textsubscript{pid}.ops[op.location]= op}
\State \Call{Propagate}{l\textsubscript{pid}.parent}
\State \Return \Call{Index}{l\textsubscript{pid}, op.location, b} \Comment we don't know the block eachoperation resides in leaf
\EndFunction{Append}

@@

\Function{void}{Propagate}{\textsl{node} n}
\If{\texttt{n {\keywordfont is} root}} \Return
\ElsIf{\texttt{not} \Call{Refresh}{n}}
\State \Call{Refresh}{n}
\EndIf
\State \Call{Propagate}{n.parent}
\EndFunction{Propagate}

@@

\Function{boolean}{Refresh}{\textsl{node} n}
\State \texttt{i= n.last}
\State \texttt{new, c\textsubscript{left}, c\textsubscript{right}= \Call{CreateBlock}{n, i+1, n.counter}}
\If{\Call{CAS}{n.blocks[i+1], null, new}}
\State \texttt{\Call{CAS}{n.last, i, i+1}}
\ForEach{\texttt{dir} {\keywordfont{in}} \texttt{\{left, right\}}}
\State \texttt{\Call{CAS}{n.dir.super[c\textsubscript{dir}], null, i+1}}
\State \texttt{\Call{CAS}{n.dir.counter, c\textsubscript{dir}, c\textsubscript{dir}+1}}
\EndFor
\State \Return{\texttt{true}}
\Else ~\Return{ \texttt{false}}
\EndIf
\EndFunction{Index}

@@

\Function{block, int, int}{CreateBlock}{\textsl{node} n, \textsl{int} i, int g} 
\Statex\Comment{Creates a block to insert into \texttt{n.blocks[i]}}
\State \texttt{block b= \Call{new}{\textsl{block}}}
\If{\texttt{n is leaf}}\Comment{TODO}
\Else
\State \texttt{b.group= g}
\ForEach{\texttt{dir} {\keywordfont{in}} \texttt{\{left, right\}}}
\State \texttt{j= n.dir.last}
\State \texttt{start= n.blocks[i-1].end\textsubscript{dir}}
\State \texttt{end= n.dir.blocks[j]}
\State \texttt{c\textsubscript{dir}= n.dir.blocks[j].group}
\State \texttt{b.end\textsubscript{dir}= j}
\State \texttt{b.num\textsubscript{dir}= end.sum\textsubscript{\#dir} - start.sum\textsubscript{\#dir}}
\State \texttt{b.sum\textsubscript{\#dir}= n.blocks[i-1].sum\textsubscript{\#dir} + b.num\textsubscript{dir}}
\EndFor
\State \texttt{b.sum= b.sum\textsubscript{left} + b.sum\textsubscript{right}}
\EndIf
\State \Return \texttt{b, c\textsubscript{left}, c\textsubscript{rightt}}
\EndFunction{CreateBlock}

@@

\begin{algorithm}
\caption{Block Tree Continued}
\begin{algorithmic}[1]

\Function{element}{Get}{\textsl{int} i} 
\State \texttt{res= \Call{BSearch}{root, sum, i, 0, root.last}}
\State \Return{\Call{Get}{root, res, i-root.blocks[res-1].sum}}
\EndFunction{Get}

@@

\Statex $\leadsto$ \textsf{Precondition: \texttt{n.blocks[start..end]} contains a block with feild \texttt{f} $\geq$ \texttt{i}}
\Function{int}{BSearch}{\textsl{node} n, \textsl{field} f, \textsl{int} i, \textsl{int} start, \textsl{int} end}

@@

\Function{element}{Get}{\textsl{node} n, \textsl{int} b, \textsl{int} i} 
\Comment{\textmd{Returns the \texttt{i}th operation in \texttt{b}th block of node \texttt{n}}}

\If{\texttt{i $\leq$ n.blocks[b].num\textsubscript{left}}} \Comment{\texttt{i} exists in left child of \texttt{n}}
\If{\texttt{n.left {\keywordfont is} leaf}} \Return \texttt{n.left.ops[i]}
\Else
\State \texttt{subBlock= \Call{BSearch}{n.left.sum, i, n.blocks[b-1].end\textsubscript{left}+1, n.blocks[b].end\textsubscript{left}}}
\State \Return\Call{Get}{n.left, subBlock, i-n.left.blocks[subBlock-1].sum} 
\EndIf
\Else
\State \texttt{i= i-n.blocks[b].num\textsubscript{left}}
\If{\texttt{n.right {\keywordfont is} leaf}} \Return \texttt{n.right.ops[i]}
\Else 
\State\texttt{subBlock=\Call{BSearch}{n.right.sum, i, n.blocks[b-1].end\textsubscript{right}, n.blocks[b].end\textsubscript{right}}}
\State \Return\Call{Get}{n.right, subBlock, i-n.right.blocks[subBlock-1].sum} \EndIf
\EndIf
\EndFunction{Get}

\Function{index}{Index}{\textsl{node} n, \textsl{int} i, \textsl{int} b} \Comment{Returns order in  the root of \texttt{i}th operation in the node \texttt{n} which exists in block b}
\If{\texttt{n {\keywordfont is} root}} \Return \texttt{i}
\Else
\State \texttt{dir= (n.parent.left==n)? left: right}
\State \texttt{superBlock= \Call{BSearch}{n.parent, n.sum\textsubscript{dir}, i, sup[n.blocks[b].group]-p, sup[n.blocks[b].group]+p}}
\If{\texttt{dir {\keywordfont is} left}}
\State \texttt{i+= n.parent.blocks[superBlock-1].sum\textsubscript{right}}
\Else \State \texttt{i+= n.parent.blocks[superBlock].sum\textsubscript{left} + n.parent.blocks[superBlock].sum\textsubscript{left} + n.blocks[nBlock-1].sum}
\EndIf
\State \Return\Call{Index}{n.parent, i, superBlock}
\EndIf
\EndFunction{Index}

@@

\section{Implementing Queue using Block Tree}
\subsection{Idea in a nutshell}
With the ideas introduced in block tree we are going to create a shared wait-free queue with \textsc{O}$(\log p)$ steps. A queue consitsts of two operations, enqueue and dequeue. \textsc{Enqueue}$(n)$ appends element $n$ to the sequence stored, this is somehow similar to \textsc{Append}$(n)$. Since \textsc{Append}$(n)$  adds $n$ to the history of all operations. \textsc{Dequeue()} returns first not dequeued element among all elements. If index $i$ is the tail of the queue, we can return the dequeue response using \textsc{GetElement}$(i)$.  So in the rest we modify block tree to compute $i$ for each \textsc{Dequeue}$()$ to achieve a FIFO queue.

\subsection{What should a \textsc{Dequeue}$()$ return, regarding history of operations?}
\paragraph{}
In a basic look of we use block tree to implement queues the block tree, we maintain the sequence of all elements(history of all operations), not only the current state of the queue. Regarding the current state of the queue, if it is not empty, the \textsc{Dequeue}$()$ returns the first existing element. And if the queue is empty it returns \texttt{null}. Now look at all the history of operations, what should $i$th \textsc{Dequeue}$()$ return?

@@

\paragraph{}
In the example above, \textsc{Dequeue}$()$ operations return \texttt{5, 2, 3, null, 4} in order. If the queue size is greater than $0$ and $r$ Dequeue operations have returned a non-null response, then $i$th Dequeue returns $r+1$th Enqueue input. So in order to answer a Dequeue, there is only needed to know the size of the queue and $\#$ returning dequeues.

\texttt{DEQ[i] = (size>0) ? ENQ[r+1] : null;}
\begin{definition}
After a sequence of operations, $\#$returning dequeues is the number of Dequeue operations that have returned non null value. \texttt{DEQ[i]} is the $i$th Dequeue. \texttt{ENQ[i]} is the argument of $i$th Enqueue.
\end{definition}

\subsection{What should a \textsc{Dequeue}$()$ return, regarding history of blocks of operations?}

\paragraph{}
In the Block Tree idea, we did not store the sequence of operations themselves but instead stored blocks of concurrent operations to optimize \textsc{Propagate()} steps and increase parallelism. So now the problem is to find \texttt{ENQ[j]} for \texttt{DEQ[i]}. From lemma \ref{lem:block_size} we know we can linearize operations in a block in any order; here, we choose to decide to put Enqueue operations in a block before Dequeue. In the next example, operations in a cell are concurrent, and the order within a block is as said before. \texttt{DEQ()} operations return \texttt{null, 5, 2, 1, 3, 4 , null} respectively.
